none

### 2024-07-04 paper/neural_2024/poster_text/results2.tex
Text for poster

% Seb to fix: Expected not experiment (in figs), Fig 8 - monochromes. Fig 1 - monochromes
% Stuart to do: Intro/Start of results modifications

\begin{document}
%\setlength{\droptitle}{-1.8cm} % move the title up a suitable amount
\maketitle

\begin{abstract}
The retinotectal projection, in which the axons of retinal ganglion cells form an ordered map in the optic tectum, is an important developmental system that has been studied for many decades. There is a large body of evidence, consisting of surgical and genetic manipulations, that give insight into how it behaves. Experimental work has uncovered a number of mechanisms that are involved in aspects of the mapping. However, a single, consistent theoretical framework has yet to describe all the patterns that are formed. Sperry's chemoaffinity theory provides an account of the wildtype pattern. Gierer formalised and extended this theory to incorporate competition to account for a subset of surgical manipulation experiments, but it was never tested against the full range of surgical manipulations. Other theoretical and computational studies have been partially successful in accounting for surgical manipulations, but to date, no single theory, operating only by local interactions, has accounted for every manipulation. We present a two-dimensional model of the retinotectal projection in which orthogonal pairs of opposing tectal ligand gradients combined with retinal gradients of receptors form Gierer's potential functions and then show that this receptor-ligand model accounts for the full range of surgical manipulations. We then extend the detail with which the receptor-ligand interactions are treated to successfully account for more recent genetic manipulation experiments.
\end{abstract}

\section{Introduction}

Since the pioneering work of Sperry \citep{sperry_reestablishment_1942,sperry_visuomotor_1943,sperry_chemoaffinity_1963}, progress on understanding topographic patterning of brain connectivity has been made through a synergy of experimental and computational neuroscience, focussed on the projection of retinal ganglion cell axons into the optic tectum. Since computers first became available as scientific tools, a wide range of theories have been developed and tested through formal algorithms and computer simulation work. Models can be distinguished on a number of grounds, such as the level of abstraction at which they describe the process of map development, and the assumptions they represent about the contributions of chemotaxis, molecular interactions between axons, and competition for space in the destination tissue (for reviews see \citealp{Prestige1975,Goodhill2005}).

One important class of theories assumes that map formation is driven by chemotaxis, influenced by pairs of anti-parallel expression gradients in the tectum, with the balance of influences in each axon determined by a receptor density specific to its origin in the retina (\citealp{Gierer1983,Gierer1987,Simpson2011}; see also \citealp{Karbowski2004,James2020}). A second class of theories assumes that a single gradient is primarily responsible for chemotaxis and that maps reflect additional processes such as competition \citep{Triplett2011}, Hebbian activity \citep{Tsigankov2006,Tsigankov2010}, or marker induction \citep{Prestige1975,Willshaw2006}.

A key test of the validity of recent models is their ability to account for data from a set of genetic manipulation studies by \cite{brown_topographic_2000} and \cite{reber_relative_2004}, in which axons from a subset of retinal ganglia cells treated with a Eph-A3 knock-in form a separate, rostrally shifted topographic map, which becomes indistinct from that formed by untreated cells, at a ‘collapse point’ on the rostral side. \cite{Sterratt2013} found that a mixture of counter-gradients and an additional competition term provided the best account of these data. A follow-up study in which a range of different computational models were systematically compared \citep{hjorth_quantitative_2015} found that \redcmnt{one focused on competition
% Stuart: I think the model is Koulakov's, not Triplett's. In Triplett2011 they give evidence for a competition model (like Koulakov's) by reducing the number of RGCs in the Math5 mutant mouse. *Hjorth* shows that the Koulakov model is the best one at reproducing the collapse in the EphA3 ki experiment.
by \cite{Triplett2011}} could best account for several such experimental findings, and that the other types of model could not explain the collapse point.

\redcmnt{The model of \cite{Triplett2011} describes maps as minima of an energy function}, whose structure can be obtained via a stochastic optimisation process. Accordingly, the energy is the sum of three terms, relating to the influences of chemical gradients, axon density (competition), and activity-dependent growth processes. Maps that optimise this function can be obtained by starting with a random map and then substituting connections at random, accepting those substitutions with a probability that increases if they reduce the energy, and repeating until the algorithm converges on a pattern of connections corresponding to a minimal energy state.

The ability of this model to explain the collapse point in the Eph-A3 knockin studies can be understood in terms of its relationship to simpler models developed previously, with which it shares (via the activity-dependent term in the energy function) similar assumptions \citep{Koulakov2004}. According to this earlier model, pairs of adjacent axons on the tectum are evaluated iteratively and the probability of their positions being switched increases with an energy defined by the product of the difference in Eph receptor density and the difference in ephrin ligand density. As the retinal receptor and tectum ligand densities express anti-parallel gradients, a positive energy is translated into a pairwise switch of axon positions with high probability. Eph knock-ins are assigned higher receptor density values, and the map collapse point emerges when differences in receptor density (due to retina position and treatment condition) are offset by differences in ephrin ligand density in the tectum. The two terms are balanced at a point that shifts further in the direction of cells with a more temporal retina origin for a stronger (i.e., homozygous) knock-in, as observed experimentally.

These models inherit from thermodynamic models used to investigate phase transitions in statistical physics, the concept of a `temperature', via a parameter that weights the switching probabilities and effectively determines the sharpness of the collapse between treated and untreated maps. Interestingly, at high temperatures (sharp switching) the model of \cite{Koulakov2004} becomes formally equivalent to the `arrow’ model, one of the first models of retinotectal map self-organisation proposed  \citep{Hope1976}.

The explanatory power of these `switching' models stems from their simplicity and degree of abstraction, revealing the map pattern that optimises the constraints represented by a given energy function. However, while switching models are evaluated through a numerical procedure (i.e., iterative switching), that procedure should not necessarily be interpreted as an explicit representation of the process of development \citep{Wilson2015}. Consider, for example, that from more realistic initial conditions than the uniform random map typically assumed, local energy minima quite unlike organised biological maps are more likely to be encountered.

In a parallel strand of work the principles represented by switching models have been explored in models that represent the developmental process more explicitly. For example, \cite{Overton1982} showed how the net effects of iterative switching in discrete systems could be achieved via a continuous dynamics based on iteratively ‘nudging’ axons by a velocity vector that is weighted by the factors considered by the arrow model, including for example, functions of the proximity of growing axons to their neighbours.

Continuing this strand of work, \cite{Simpson2011} have more recently extended the model of \cite{Overton1982} from its original description of axon growth in one spatial dimension to simulate growth in two dimensions, and they have shown that it can account for a wide range of data from genetic studies, as well as a wide range of surgical manipulation experiments; including the emergence of a map collapse point in simulated Eph-A3 knock-in experiments.

The models of \cite{Overton1982} and \cite{Simpson2011} incorporate three basic ingredients; competition for space, receptor-based interactions between axons, and gradient-based guidance. However, a major limitation is that chemotaxis is not implemented explicitly. Instead, simulations involve providing growing axons with explicit directional information about their target destination at every timestep. This was an appropriate choice given the scientific goals of \cite{Simpson2011}, but for this reason the theory of retintopic map formation that is represented by their model cannot be tested comprehensively, and indeed this model was excluded from the systematic comparison of models by \cite{Hjorth2015} on that basis.

An open question then, which we address in this paper, is whether the \cite{Simpson2011} model can still account for a wide range of experimental data from surgical and genetic experiments, when its assumptions about chemotaxis are represented explicitly, i.e., when simulated axon growth is guided by only local information? If so, such a model would represent the most complete account of retinotectal map self-organization to date.

** Maybe we also need to say that a second objective to to express the model
exclusively in terms of receptor-ligand interactions.

%Sperry's chemoaffinity model is an influential theory of how axons from neurons that originate in one tissue can find their way to specific locations in a target tissue during early brain development \citep{sperry_reestablishment_1942,sperry_visuomotor_1943,sperry_chemoaffinity_1963}.

%The central proposal of the theory is that growing axons find their location in a destination tissue by means of `labels of a cytochemical nature'. The chemical affinity between labels in source and destination tissues allows axons to distinguish correct target locations from incorrect ones.

%This key idea (called chemospecificity) does not, on its own, explain how an axon can find the correct \emph{direction} in which it should move to reach its target. An additional component of Sperry's theory \citep{stevens_handbook_1950,sperry_problems_1955,sperry_chemoaffinity_1963} suggested that the axon could determine this direction by evaluating local information in gradients of chemicals spanning the target tissue, giving rise to chemotaxis of the axonal growth cones.

%Much of Sperry's work was carried out on the retinotectal system, which is ideal for the study of ordered axonal development in fish, amphibia and birds, because in these animals, patterning of axons across the tectum is largely complete prior to the onset of neural activity \citep{oleary_molecular_1999}.

%In this system a two-dimensional map of  the retina is developed on the surface of the optic tectum such that the axons from adjacent retinal ganglion cells (RGCs) find their way to adjacent locations on the tectum.  This retinotectal map is amenable to a number of surgical manipulations especially in amphibia, in which re-growth of the projection can be observed after surgical changes to the sensory apparatus (such as ablation or rotation of the retina) or the map substrate (the tectum). This allowed Sperry and other researchers to carry out a wide range of experimental manipulations, the results of which informed the chemoaffinity theory.

%In the 1990s, the theory of chemoaffinity and chemotaxis was given robust support by the discovery of the ephrin ligands and their receptors \citep{cheng_complementary_1995,drescher_vitro_1995} which have been shown to form into graded expression fields in the retina \citep{braisted_graded_1997}, tectum \citep{braisted_graded_1997,feldheim_genetic_2000} as well as in other sensory systems, such as the somatosensory system \citep{vanderhaeghen_mapping_2000}. % And sounds?

%The Ephrin ligands have a clear effect on axonal outgrowth, as shown in-vitro \citep{cheng_complementary_1995,drescher_vitro_1995,hansen_retinal_2004} and by in-vivo studies \citep{frisen_ephrin-a5_1998,rodger_transient_2000,mann_topographic_2002,hindges_ephb_2002}. Two classes of ephrins have been identified, labelled `A' and `B'. Ephrin A and its receptors mediate a repulsive interaction which contributes to an ordering of RGC axons on the tectum along the rostral-caudal axis. Ephrin B acts on medial-lateral ordering with an attractive interaction. Together, the ephrins are able to provide a coordinate system on the tectum to guide growing axons, but precisely how this positional information is used by axonal growth cones is not yet fully understood.

%Quantitative models exploring axonal ordering were already in existence by the time of the discovery of the ephrins.  The importance of chemical gradients to biological processes was already well known from the study of slime moulds \citep{bonner_evidence_1947}.  Sperry himself discussed chemical gradients in relation to the problem of neuronal pathfinding \citep{sperry_problems_1955}, albeit qualitatively. Some time after Sperry's active period, but before the ephrins were identified, \citet{gierer_development_1981,gierer_model_1983,gierer_directional_1987} made one of the few quantitative investigations of a gradient-based model of the retinotectal projection. He presented gradient following as a special case of a general class of \emph{potential models}. In a potential model, the interactions between the properties of a growing axon originating from a location $\mathbf{u}$ on the retina and its environment at location $\mathbf{x}$ on the tectum set up some potential function, $p(\mathbf{x},\mathbf{u})$.  The axon then grows or moves to minimize $p$, which causes it to find its target location. This model successfully creates the wildtype map because each axon originates with a different $\mathbf{u}$ and has its minimum $p(\mathbf{x},\mathbf{u})$ in the correct target location. This model forms a mathematical expression of Sperry's theory. Because it is quantitative, it allowed Gierer to identify two features of the theory: i) A \emph{pair} of opposing gradients is needed to form a potential function in one dimension (and \emph{two} pairs to create a 2D potential function). ii) Tectal or retinal ablation would require $p$ to be re-specified to reproduce the experimental results that fibres from an ablated retina will generate a map across the entire tectum and fibres from an intact retina will form a compressed map on an ablated tectum \citep{attardi_preferential_1963,schmidt_retinal_1978,schmidt_expansion_1978}.

%Gierer addressed the second problem by introducing a form of competition between fibres, postulating that $p$ would accumulate locally at a rate proportional to the fibre density in that location and he noted that this was equivalent to a more direct fibre-fibre competition. He demonstrated, in a one-dimensional computer simulation, that this form of competition would lead to the expected response to ablations \citep{gierer_model_1983}.

%Gierer's simulations confirmed the validity of this basic idea, which we will refer to as \emph{differential chemotaxis}, for the problem of map formation with respect to a single spatial dimension \citep{gierer_model_1983,karbowski_model_2004}.  The model helped establish how \emph{competition between axons} for space in the target tissue can additionally account for  map development after tissue ablation.

%Simulations of differential chemotaxis and axonal competition have recently been extended to investigate map development in two spatial dimensions, revealing how these two ingredients can give rise to realistic patterns of synaptic innervation \citep{james_modelling_2020}.  These models are formulated as systems of partial differential equations (PDE), and solved through numerical integration on an underlying lattice structure representing the target tissue.  In general, this approach makes mathematical analysis of system dynamics tractable, and as such PDE modelling has played a central r\^ole in the development of theories of pattern formation (see \citet{krause_modern_2021} for recent examples).  However, we found it numerically intractable to extend the model described in \citet{james_modelling_2020}, which simulated 41 whisker-specific projections, to the retinotectal system for which hundreds of projections are required (to demonstrate surgical manipulations). Instead, we sought to implement an \emph{agent-based} model of differential chemotaxis, in which each axon growth cone is an agent that interacts with its environment and with the other growth cone agents in the system. With this model, we asked: i) Can a two-dimensional implementation of Gierer's model explain the full range of surgical manipulations that have been described in the literature? ii) Can the model be developed such that only \emph{local} interactions are required to transfer information between agents? and iii) Can a differential chemotaxis model explain more recent results in which levels of EphA receptor sub-types are changed by genetic manipulation \citep{brown_topographic_2000,reber_relative_2004}?

\section{Results}

%% * receptors are expressed by RGCs as f(x,y)=... (Eq. 1)

Each of $n$ retinal ganglion cells (RGC) projects $M$ growth cones (referred to as \emph{branches}) which carry a set of $N$ different receptor types, expressed at levels determined by the cell soma's location on the retinal grid. RGC growth cones and cells on the tectum are also assumed to express $N$ different ligand types which interact with RGC receptors.
%
We adopt the same form for the expression of receptors by retinal ganglion cells as
\citet{simpson_simple_2011}:
\begin{equation} \label{e:f}
  f(x,y) = 1.05 + 0.26 \exp(2.3 u),
\end{equation}
where $u$ gives the direction with which the expression increases and may be substituted by $(1-x)$, $(1-y)$, $x$ or $y$. We defined $x$ as an axis from temporal to the nasal retina and $y$ from ventral to dorsal retina.

%% * Simpson and Goodhill modelled according to deltax = B + C + G (Eq 2)
In Simpson \& GoodHill's model of chemotaxis and competition, the position, $\mathbf{x}_t$, of each branch $b$ on a simulated tectum can be updated according to $\mathbf{x}_{t+1} = \mathbf{x}_{t} + \Delta \mathbf{x}$ with the movement vector given by
% Maybe re-order the terms?
\begin{equation} \label{e:dX}
 \Delta \mathbf{x} = \mathbf{B} + \mathbf{C} + \mathbf{G}.
\end{equation}
$\mathbf{B}$ is a movement applied close to the tissue boundary that ensures all branches remain within the tissue (\citet{holt_target_1998}---see method details).
%
%% * They had C composed of two terms, distance based and a component based on EphA expression
%
$\mathbf{C}$ is the movement due to competitive interactions with other branches
%
and $\mathbf{G}$ is the movement vector given rise to by the chemotaxis effect.

%% * To represent the mechanism for comp. explicitly, we introduced receptor AND ligand expressions

%%We reformulated this so that the distance based repulsion is also mediated via receptor ligand interactions. % No need to justify yet - can go in disucssion
%

%%%%%%%% Remove/coment out %%%%%%%%%%%%%%%%%
% No need for this rhetorical stuff
%\citet{simpson_simple_2011} investigated a model of this form in which a `placeholder' provided a chemotaxis-like movement for $\mathbf{G}$ and a form for $\mathbf{C}$ was studied.

%Their model, which successfully accounts for a wide range of experimental
%manipulations, foundxxx *Their model had a* $\mathbf{C}$ comprised of two components; a simple distance-based competition between \emph{all} branches, along with an axon-axon interaction based on the relative levels of the receptor EphA expressed by each interacting axon.
% Rewrite next little bit
% Say the global supervisor comes into the expression G, but having G a vector
% that always points towards...
% This may come later

%We asked i) if the placeholder mechanism for $\mathbf{G}$, in which a `global supervisor' informs each branch of the vector along which it should travel to approach its target, could be replaced by a true chemotaxis mechanism based on graded signalling molecule expression and ii) if the competition mechanism found by Simpson and Goodhill was the simplest possible mechanism that could explain all of the surgical and genetic manipulations accounted for by their model.
%%%%%%%%%%%%%%%%%%%%%%%

%  Maybe lose the sub sub sections
\subsubsection*{A signalling model for competition}

Simpson \& Goodhill's model had a $\mathbf{C}$ comprised of two components; a simple distance-based competition between \emph{all} branches, along with an axon-axon interaction based on the relative levels of the receptor EphA expressed by each interacting axon.
%Although the model incorporated EphA expression, the distance-based competitive interaction was given no mechanism of operation.
To provide an explicit mechanism of operation for the competition, we modelled it as a sum of interactions that are `switched on' via receptor-ligand signalling, meaning that some axons would have no competitive interaction, while others would repel strongly.
This single-component competitive movement vector for a branch $b$ is

\begin{equation} \label{e:X}
\mathbf{C} = \frac{m_{\!_X}}{|B_{b}|} \sum_k^{nM} \hat{\mathbf{x}}_{kb}\,W\,Q(d_{kb}, \mathbf{r}_{b}, \mathbf{l}_{k}).
\end{equation}
%
$m_{\!_X}$ is a scalar parameter and $B_{b}$ is the set of branches that are within interaction distance ($2 r_{\!_X}$) of $b$. $r_{\!_X}$ is the receptor-ligand interaction radius. $nM$ is the total number of branches in the system and $\hat{\mathbf{x}}_{kb}$ is the unit vector from branch $k$ to $b$.
%
$W$ is the distance based weighting ($W = 1-\frac{d_{kb}}{2r_{\!_X}}~\mathrm{if}~  d_{kb}\leq 2r_{\!_X}$, otherwise $0$).
%
The distance-based weighting can be interpreted as being related to the probability with which a ligand may bind to a receptor and mediate the interaction.
%
$Q$ is a signalling threshold function which depends on the distance $d_{kb}$ between branches $b$ and $k$, the receptor expression on $b$ ($\mathbf{r}_b$) and the ligand expression on $k$ ($\mathbf{l}_k$).
$Q$ is set to 1 only if at least one repulsive signal exceeds a positive signal threshold, $s$, and the distance from branch $b$ to branch $k$ is smaller than $2 r_{\!_X}$:
%
\begin{equation}
Q(d_{kb}, \mathbf{r}_{b}, \mathbf{l}_{k}) = \begin{cases}
                 0 & \mathrm{if}~-F_i\,r_{i,b}\,l_{i,k} <
                 s,\,\forall{i}~\mathrm{or}~d_{kb} > 2r_{\!_X} \\
                 1 & \mathrm{otherwise.}
     \end{cases}
\end{equation}
%
$l_{i,k}$ (the expression of ligand $i$ on branch $k$) is the $i^{\mathrm{th}}$ element of $\mathbf{l}_k$ and $r_{i,b}$ is the $i^{\mathrm{th}}$ element of $\mathbf{r}_b$.
$F_i$, defined in Eq.\,\ref{e:G}, is -1 if the $r_{i}\,l_{i}$ interaction signals repulsion and +1 if it signals attraction.
With the constraint $s>0$ comes the assumption that attractive interactions will have no effect on the competitive movement vector---a receptor-ligand pair that signals attraction cannot cause $Q$ to take the value 1.

\subsubsection*{A gradient based model of chemotaxis}

% Say the global supervisor comes into the expression G, but having G a vector
% that always points towards...
% This may come later

@@

Each of the $M$ RGC growth cones carry a set of $N$ receptors, $r_i$, indexed by $i$ and expressed at levels determined by the cell soma's location on the retinal grid.
%
%The expression of each receptor varies only with respect to one spatial dimension.
%
We defined 4 receptor expression gradients (each varying only wrt one spatial dimension) arranged in orthogonal pairs with the gradient of $r_0$ being orthogonal to that of $r_1$. $r_2$, whose gradient is opposite to $r_0$, was orthogonal to $r_3$ (Fig.\,\ref{f:ex}B). Thus the model is of \emph{differential chemotaxis}.

% NO NEED FOR This sentence/bit (keep comment in case for discussion)
%Because there is convincing evidence that EphA and EphB receptors are expressed in exponentially increasing patterns \citep{reber_relative_2004,feldheim_genetic_2000,brown_topographic_2000,koulakov_stochastic_2004}, we use an exponential form for retinal receptor expressions, in common with other modelling studies~\citep{reber_relative_2004,koulakov_stochastic_2004,simpson_simple_2011}.
%
% Already said this:
%We adopt the same precise form for the retinal receptor expression as \citet{simpson_simple_2011}, $f(x,y) = 1.05 + 0.26 \exp(2.3 u)$, where $u$ gives the direction with which the expression increases and is substituted by $(1-x)$ for $r_0$, $(1-y)$ for $r_1$, $x$ for $r_2$ and by $y$ for $r_3$ (Fig.\,\ref{f:ex}).

Cells on the tectum express ligands, $L_i$, for the retinal receptors, also in orthogonal pairs of gradients.
Although several studies model tectal ligand expression with exponential functions \citep{koulakov_stochastic_2004}, the experimental evidence for the form of tectal ligand expression is more ambiguous than that for retinal receptor expression (For example data, see Fig.\,\ref{f:expr_ephb} which shows expression patterns for EphB/ephrinB).
We kept in mind the possibility that tectal ligand expression may be modelled by another function, but initially, we set it to the same exponential used for retinal expression and used the substitutions $y$, $x$, $(1-y)$ or $(1-x)$ in place of $u$ to obtain the ligand expression functions for $L_0$, $L_1$, $L_2$ and $L_3$ (Fig.\,\ref{f:ex}).

%While we do not explicitly name $r_0$, $r_1$, etc.~as \emph{EphA}, \emph{EphB}, the suggestion is that $r$ could include EphA, EphB, Ryk \citep{schmitt_wntryk_2006} and Neogenin \citep{rajagopalan_neogenin_2004} receptors and that $L$ could include the ephrin-A, ephrin-B, Wnt3 \citep{schmitt_wntryk_2006} and RGM \citep{monnier_rgm_2002} ligands, each of which has been shown to play a r\^ole in retinotectal map formation.
%%%%%%%%%%%%%%%%%

\subsubsection*{Simulation}

The model has four free parameters; $m_{\!_G}$, $m_{\!_X}$, $r_{\!_X}$ and $s$. We fixed $m_{\!_G}$ to the value $0.0012$ and used a numerical optimization process to find the competition parameters that would balance this level of chemotaxis. The parameters found were $m_{\!_X} = 0.2078$, $r_{\!_X} = 0.0759$ and $s = 4.815$ (see method details).

% Results proper here Leave this as is for now
\subsection*{Retinotectal map self-organization}

Fig.\,\ref{f:GJ} shows the simulation results with these parameters after 1500 time steps.
Fig.\,\ref{f:GJ}A indicates the colour code used for retinal cell position; dorsal retinal cells are indicated by green, nasal cells by red. Lightness gives position along the relevant axis so that cells at the dorsal-nasal corner are yellow and those at the ventral-temporal corner are black.
Fig.\,\ref{f:GJ}E shows the expected layout of retinal ganglion cell axons on the tectum.
The axons retain the colour of their originating cell on the retina.
Nasal cell axons should be arranged caudally; dorsal axons laterally; axons should be evenly spaced and the pattern should span the entire tectum.
Fig.\,\ref{f:GJ}, panels B--D show `fishnet' plots of the average location of the 4 branches of each axon (the axon's centroid) at three timepoints.
Fig.\,\ref{f:GJ}B indicates the randomised placement of axon branches at the rostral end of the tectum as specified by the initial conditions (Eq.\,\ref{e:ic}).
In these plots, lines are drawn between axons that are expected to arrange themselves adjacently to one another by the end of the simulation.
The number of line crossings in this plot gives an indication of the disorder in the arrangement.
Fig.\,\ref{f:GJ}G shows the positions of the individual branches (i.e.~growth cones) at $t=1500$ that contribute to the plot in Fig.\,\ref{f:GJ}D.
By this time, the topological order matches that shown in Fig.\,\ref{f:GJ}E.
Opposing gradients have sorted the axons along the two axes to form a grid of cells.
The arrangement spans the full tectum due to competition between the growth cones (in the chemotaxis-only model with $m_{\!_X}=0$, axons cover only a fraction of the tectal tissue).
Fig.\,\ref{f:GJ}F shows the two pattern metrics; $\epsilon$ (distance error) and $\eta$ (number of crossings).
For the wildtype pattern, $\eta_t=0$, as in Fig.\,\ref{f:GJ}E.
Fig.\,\ref{f:GJ}H shows selected axons (one from the centre and one from each corner of the retina) and their position history.

% Stuart to update the text here
\subsection*{Surgical manipulations}

After finding an optimum set of parameters for the wildtype arrangement, we
tested the model with a number of simulations of surgical manipulations
%(for reviews, see \citet{udin_formation_1988} and
%\citet{goodhill_retinotectal_1999})
.
Figs.~\ref{f:GJsurg_grafts}A and \ref{f:GJsurg_grafts}B show the results of tectal graft rotation manipulations in which a square patch of tectal tissue is cut out and rotated $90\degree$ or $180\degree$ \citep{chung_observations_1978}.
%
To simulate the surgery, we rotated an $8\times8$ square of the tectum shown in Fig\,\ref{f:ex}D, carrying the ligand expressions with the rotation (resulting in Fig.\,\ref{f:trot90}, left column).
After the manipulation, we re-computed the tectal ligand expression gradients. Fig.\,\ref{f:trot90} demonstrates that the most significant change in the gradient is around the border of the graft, due to discontinuities in the ligand expression.
The manipulated gradients, interacting with unchanged axon branches, led to a qualitative rotation of axons that were destined for the graft area, indicated by a 90$\degree$ or 180$\degree$ `twist' of the axon centroid colours.
Long connecting lines in the fishnet show that a number of axons from near the edges of the tectum were drawn into the graft region.
The pattern metric, $\epsilon$, is shown in the right hand graphs of Fig.\,\ref{f:GJsurg_grafts}A \& B.
We plot $\epsilon$ (the distance error) for the entire map (blue) along with the error for the graft region (gold) and the surround region (red).
The overall distance error for the rotation manipulations (0.182 or 0.2752)  is greater than for the wildtype system (0.04).
The pattern within the graft exhibits greater average distance error than for the surrounding tissue.
The system does not achieve a perfect reproduction of the target layout suggested by the experimental result (which is shown in the inset) but it is in qualitative agreement.
%
Fig.\,\ref{f:GJsurg_grafts}C shows the result of a similar manipulation in which two grafts are cut out of the tectum and swapped.
There is considerable disorder in the simulated pattern of axon centroids but a line of red/yellow agents that would been located caudally has advanced rostrally and a number of green agents are more caudal than in the wildtype arrangement, consistent with experimental observations.

Fig.\,\ref{f:GJsurg_ablate}A shows the result of a retinal ablation, in which the temporal half of the retina is surgically removed and the remaining RGC axons are allowed to develop on the unmanipulated tectum.
According to a pure chemoaffinity theory, the surviving retinal axons should retain their chemical labels and they would be expected to find their usual locations on the tectum, resulting in a tectum which is only half populated with axons.
The experimental observation is that the axons arrange themselves across the whole tectum, retaining their topographic order (inset, \citet{attardi_preferential_1963,schmidt_expansion_1978}).
The fishnet plot of the simulation indicates that the competition in the system successfully `stretches out' the surviving axons to achieve a close representation of the experimentally observed result.
In this figure, the right-hand graph shows the simulated position of each axon centroid on the rostro-caudal (R--C) tectal axis plotted against its position on the nasal-temporal (N--T) retinal axis in red triangles, with the experimental positions plotted in blue squares.
%
Fig.\,\ref{f:GJsurg_ablate}B shows the result of a tectal ablation. Here the retina is left un-modified, but the caudal half of the tectum is surgically removed. Experimental observation \citep{yoon_reorganization_1971,sharma_reformation_1972} shows that RGC axons `crush up' on the remaining rostral section, retaining topographic order and creating a complete map on the available tectal tissue. The model clearly replicates this result, indicated by both fishnet and tectal/retinal axes graphs, because the chemotaxis component of the model is based on balanced, relative interactions between opposing gradients, rather than absolute labels or expression measurements.
%
A related ablation experiment is the so-called `mismatch', in which the temporal half of the retina is ablated along with the caudal part of the tectum to which it normally maps, again challenging the pure chemoaffinity model \citep{horder_retention_1971}. The competitive differential chemotaxis model replicates this manipulation (Fig.\,\ref{f:GJsurg_ablate}C) for the same reason as in the simple tectal ablation of Fig.\,\ref{f:GJsurg_ablate}B; opposing gradients remain on the surviving tectum allowing RGC axons to find their relative ordering.
%
The final surgical manipulation which we investigated was the `compound eye' experiment, in which two nasal retina halves are fused together \citep{gaze_retino-tectal_1963,fawcett_retinotectal_1982}. Within the compound retina there exist matched pairs of retinal ganglion cells, one from each nasal half, whose receptor and ligand expressions cannot be distinguished. These form overlaid maps, each of which spans the entire tectum (Fig.\,\ref{f:GJsurg_ablate}D).

% Stuart to work on...
\subsection*{Genetic manipulations}

% Context that might need to go in the Discussion
%In the early 2000s, a series of genetic experiments added to the set of
%manipulations that a model of the retinotectal system would need to explain.

% Modify following so that the discursive stuff goes in the discussion.
First, \citet{brown_topographic_2000} showed that a selective RGC knock-in of EphA3 receptors led to an interesting modulation of the maps found in the tectum.
For the strongest knock-in (a homozygote manipulation) two maps were formed, with knock-in cells finding termination zones more rostral than their wildtype retinal neighbours.
By enhancing the quantity of EphA receptors expressed by half of all retinotectal axons, they demonstrated the importance of retinal receptor expression for the arrangement of the tectal map.
In the weaker, heterozygote manipulation, the maps were still separated, but the separation `collapsed' towards the rostral end of the tectum (see \citet{brown_topographic_2000}, Fig\,5).
This led to a second study showing that with an additional knock-down of EphA4 receptors in \emph{all} RGCs, the retinal origin of the collapse point was shifted temporally and all heterozygote EphA3 knock-in cells were pushed even further rostrally (\citet{reber_relative_2004}, their Fig.\,3e).
For the model to successfully reproduce these results, it should therefore: i)~Demonstrate a complete separation of maps for a `large' knock-in of EphA3; ii)~Show a separation of maps for a smaller knock-in of EphA3 that is greater for axons of nasal origin. iii)~For the smaller knock-in, a collapse point at 76\% of the N--T axis; iv)~Show a considerable rostral shift of cells for which EphA3 is knocked in and EphA4 is knocked down; and v)~Show a collapse point for the knock-in/knock-down cells and knock-down only cells at around 88\% of the N--T axis.

An assumption made in these papers is that all EphA receptor sub-types contribute to a summed EphA expression level, $\sum\mathrm{EphA}$, which increases from a low level at the nasal side of the retina, to a high level at the temporal side.
The precise contribution of each sub-type to the overall level is assumed not to be important.
In this model, receptor $r_0$ increases in the retina in a nasal to temporal direction (Fig.\,\ref{f:ex}B) and so we treated $r_0$ as being equivalent to $\sum\mathrm{EphA}$ and manipulated it to attempt a recreation of the \citet{brown_topographic_2000}~and \citet{reber_relative_2004}~experiments.
In half of the RGCs, we knocked EphA3 in by adding to the value of $r_0$ in these cells either 0.8 (heterozygote EphA3 ki/+) or 3.2 (homozygote EphA3 ki/ki).
To include the additional heterozygote knock-down of EphA4 (EphA4 +/-) from \citet{reber_relative_2004}, we subtracted 1.2 from $r_0$ in \emph{all} RGCs.
Fig.\,\ref{f:GJeph}A shows, for the single EphA3 knock-in (EphA3 ki/+), branch positions, a plot of axon centroids and final rostral-caudal position on the tectum versus the originating nasal-temporal retinal coordinate for time step 1500.
The left panel shows branches coloured in the red/green duochrome scheme according to their origin location on the retina.
In the middle panel, axon centroids of RGCs which were treated with the knock-in manipulation are coloured red, wildtype cells are blue.
The right hand panel mapping final R--C position versus N--T origin  (and its inset which shows the experimental result) follow the same red/blue colour scheme.
The map of treated cells (red) was shifted slightly rostrally, and the gap between the red and blue curves became smaller for more temporal RGCs. However, it can't be described as a map collapse (the term used in \citet{brown_topographic_2000}) because the curves remain separated for all N--T values. Regardless of the value of the ratio between treated and untreated cells, treated temporal axons are drawn to a significantly more rostral position than those without the knock-in. A bootstrapped studentized t-test of the two sets of points indicates a low probability, $p<10^{-4}$, that any of them were drawn from statistically identical distributions.

Fig.\,\ref{f:GJeph}B shows the same trio of graphs for the homozygote double knock-in, EphA3 ki/ki. Here, the treated axons were pushed further towards the rostral end of the tectum and the untreated axons were pushed caudally due to competition with the treated cells.
The maps were completely separated. As in the experiment, there was no collapse point.
The rostral movement of treated axons occurred due to their increased expression of EphA ($r_0$), which increased the simulated interaction with the rostral-caudal tectal ligand gradient.
The opposing receptor expression ($r_2$) was unchanged by the manipulation and so those axons with increased $r_0$ expression were pulled to a more rostral position than they would otherwise have taken.
The caudal movement of wildtype axons, whose receptor expression in Figs.\,\ref{f:GJeph}A-B was unchanged, occurred due to competition with the most caudal of the rostrally-drawn treated axons.
Overall, the correspondence between the experimental result and the simulated homozygous knock-in is close.

>>>>

\begin{document}

\subsection*{A gradient based model of chemotaxis}

@@

Each of the $M$ RGC growth cones carry a set of $N$ receptors, $r_i$, indexed by $i$ and expressed at levels determined by the cell soma's location on the retinal grid.
%
We defined 4 receptor expression gradients (each varying only wrt one spatial dimension) arranged in orthogonal pairs with the gradient of $r_0$ being orthogonal to that of $r_1$. $r_2$, whose gradient is opposite to $r_0$, was orthogonal to $r_3$ (Fig.\,1B). Thus the model is of \emph{differential chemotaxis}.

Cells on the tectum express ligands, $L_i$, for the retinal receptors, also in orthogonal pairs of gradients.
Although several studies model tectal ligand expression with exponential
functions \citep{koulakov_stochastic_2004}, the experimental evidence for the
form of tectal ligand expression is more ambiguous than that for retinal
receptor expression (For example data, see Fig.\,9 which shows expression patterns for EphB/ephrinB).
We kept in mind the possibility that tectal ligand expression may be modelled
by another function, but initially, we set it to the same exponential used for
retinal expression and used the substitutions $y$, $x$, $(1-y)$ or $(1-x)$ in
place of $u$ to obtain the ligand expression functions for $L_0$, $L_1$, $L_2$
and $L_3$ (Fig.\,1).
